Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 book/sources/intro.tex
Still cool...

A very specific type of information systems are database systems, which are computer systems designed to manage databases. A database contains data representing real objects or concepts. Each (atomic) part of these data is a result value of a measurement of a property or a description of a property of a real object or concept. In reality, some aspects or properties of objects or concepts are time-variant or time-related. E.g., the moment of a bank transaction is traditionally a moment in time and thus a time-related notion, the function of an employee in a company can change through recorded history and is thus time-variant. A temporal database schema is a database schema that models real objects or concepts with time-related or time-variant properties. However, the modelling of temporal aspects has a direct impact on the consistency of the temporal database, because the temporal nature of these aspects imposes extra integrity constraints. An example. Consider a relation in a relational library database, modelling the presence of books in the library. Every physical book is represented by a unique identifier. Every record in the relation contains such an identifier, a date on which the corresponding book was loaned and a date on which it was subsequently returned (if it was returned). Without further precautions, a library employee could add several records with the same book identifier, different `loaned'-dates and no `returned'-dates. This group of records would represent the same physical book being loaned several times on different dates and never returned, which is of course impossible. A temporal database model will typically constrain record insertion and prevent similar modelling inconsistencies.

@@

An interesting issue in temporal modelling concerns relationships between temporal notions. Notably, Allen~\cite{Allen83} studied temporal relationships between time intervals (and as a special case time points). Among others, the querying of temporal databases has greatly profited from these temporal relationships, because they allowed for richer and more complex user-specified temporal query demands, by allowing to express more complex relationships between the temporal notions in the temporal expressions in the query and the temporal indications in the database. E.g., in a relation modelling who was department head of an institution during which periods of time, a query like `who were the department heads when Thomas worked for the institution' can be evaluated using similar relationships.

Humans handle temporal information using certain temporal notions like time intervals or time points~\cite{Dyreson1994}, and they often have to deal with imperfections like imprecisions, vaguenesses, uncertainties or inconsistencies possibly contained in the descriptions of these temporal notions. Among many others, these possible imperfections in descriptions of temporal notions determine an important issue in temporal modelling. E.g., the description of the temporal notion in a sentence like `The Belfry of Bruges was finished on a day somewhere between 01/01/1201 A.D. and 31/12/1300 A.D.' contains imperfection because of the uncertainty in the used time-related expression. It is known that the building was finished on a single day, but it is not known precisely which day this was.

@@

Next to temporal modelling, some attention has been spent on temporal reasoning~\cite{Allen83}. Although temporal reasoning is not discussed in this chapter, it should be noted that, among others, Dubois and Prade~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},~\cite{Dubois89} have dealt with fuzziness and uncertainty in temporal reasoning.

The aim of this chapter is to present and explain some main concepts regarding time in information systems and to present and discuss some issues and techniques concerned with handling data imperfections related to time. The rest of this chapter is structured as follows. Section \ref{sec:time-domain} presents some basic concepts and terminology about temporal modelling and discusses some of its important aspects and issues, while section~\ref{sec:fuzzy-time-mod} presents some important issues concerning the combination of data imperfections and temporal modelling. In section \ref{sec:temporal-databases}, some basic concepts and terminology about temporal databases are presented, followed by an overview of some interesting issues concerning temporal databases and a survey of some commercial temporal database systems. Finally, in section \ref{sec:imprecision-uncertainty}, an approach to querying temporal databases containing imperfect temporal information is presented, followed by some conclusions and some suggestions for future work in section \ref{sec:further-research-work}.

@@

%Another main issue in temporal databases is the querying. The user specifies in the query some temporal constraints, e.g. \emph{``before this year''},\emph{``after 1995''}. In order to solve these temporal comparisons, between the time specified in the query and the temporal data stored in the database, some basic relationships have been defined. Allen~\cite{Allen83} first introduced the temporal relations between time intervals and, to a lesser extent, to time points. Some authors~\cite{ohlbach2004},\cite{nagypal2003},\cite{schockaert08} have softened the temporal operators defined by Allen. Therefore it is possible to compute the relationships between two ill-known time points or intervals. In~\cite{garrido2009}, some different temporal operators are defined by a combination of regular fuzzy comparisons. %e.g. the \emph{BEFORE} operator is defined by means of the fuzzy less than (FLT) operator.

>>>>

A very specific type of information systems are database systems, which are computer systems designed to manage databases. A database contains data representing real objects or concepts. Each (atomic) part of these data is a result value of a measurement of a property or a description of a property of a real object or concept. In reality, some aspects or properties of objects or concepts are time-variant or time-related. e.g., the moment of a bank transaction is traditionally a moment in time and thus a time-related notion, the function of an employee in a company can change through recorded history and is thus time-variant. A temporal database schema is a database schema that models real objects or concepts with time-related or time-variant properties. However, the modelling of temporal aspects has a direct impact on the consistency of the temporal database, because the temporal nature of these aspects imposes extra integrity constraints. An example. Consider a relation in a relational library database, modelling the presence of books in the library. Every physical book is represented by a unique identifier. Every record in the relation contains such an identifier, a date on which the corresponding book was loaned and a date on which it was subsequently returned (if it was returned). Without further precautions, a library employee could add several records with the same book identifier, different `loaned'-dates and no `returned'-dates. This group of records would represent the same physical book being loaned several times on different dates and never returned, which is of course impossible. A temporal database model will typically constrain record insertion and prevent similar modelling inconsistencies.

@@

An interesting issue in temporal modelling concerns relationships between temporal notions. Notably, Allen~\cite{Allen83} studied temporal relationships between time intervals (and as a special case time points). Among others, the querying of temporal databases has greatly profited from these temporal relationships, because they allowed for richer and more complex user-specified temporal query demands, by allowing to express more complex relationships between the temporal notions in the temporal expressions in the query and the temporal indications in the database. e.g., in a relation modelling who was department head of an institution during which periods of time, a query like `who were the department heads when Thomas worked for the institution' can be evaluated using similar relationships.

Humans handle temporal information using certain temporal notions like time intervals or time points~\cite{Dyreson1994}, and they often have to deal with imperfections like imprecisions, vaguenesses, uncertainties or inconsistencies possibly contained in the descriptions of these temporal notions. Among many others, these possible imperfections in descriptions of temporal notions determine an important issue in temporal modelling. e.g., the description of the temporal notion in a sentence like `The Belfry of Bruges was finished on a day somewhere between 01/01/1201 A.D. and 31/12/1300 A.D.' contains imperfection because of the uncertainty in the used time-related expression. It is known that the building was finished on a single day, but it is not known precisely which day this was.

@@

Next to temporal modelling, some attention has been spent on temporal reasoning~\cite{Allen83}. Although temporal reasoning is not discussed in this chapter, it should be noted that, among others, Dubois and Prade et al.~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},~\cite{Dubois89} have dealt with fuzziness and uncertainty in temporal reasoning.

The aim of this chapter is to present and explain some main concepts regarding time in information systems and to present and discuss some issues and techniques concerned with handling data imperfections related to time. The rest of this chapter is structured as follows. Section \ref{sec:time-domain} presents some basic concepts and terminology about temporal modelling and discusses some of its important aspects and issues, while section~\ref{sec:fuzzy-time-mod} presents some important issues concerning the combination of data imperfections and temporal modelling. In section~\ref{sec:temporal-databases}, some basic concepts and terminology about temporal databases are presented, followed by an overview of some interesting issues concerning temporal databases and a survey of some commercial temporal database systems. Finally, in section~\ref{sec:imprecision-uncertainty}, an approach to querying temporal databases containing imperfect temporal information is presented, followed by some conclusions and some suggestions for future work in section~\ref{sec:further-research-work}.

@@

%Another main issue in temporal databases is the querying. The user specifies in the query some temporal constraints, e.g., \emph{``before this year''},\emph{``after 1995''}. In order to solve these temporal comparisons, between the time specified in the query and the temporal data stored in the database, some basic relationships have been defined. Allen~\cite{Allen83} first introduced the temporal relations between time intervals and, to a lesser extent, to time points. Some authors~\cite{ohlbach2004},\cite{nagypal2003},\cite{schockaert08} have softened the temporal operators defined by Allen. Therefore it is possible to compute the relationships between two ill-known time points or intervals. In~\cite{garrido2009}, some different temporal operators are defined by a combination of regular fuzzy comparisons. %e.g., the \emph{BEFORE} operator is defined by means of the fuzzy less than (FLT) operator.

### 2012-03-16 book/sources/timeDomain.tex
Still cool...

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in \cite{Bohlen1998}.

@@

\begin{example}
\label{examplegrangraph}
Consider a system that models both Gregorian calendar dates as well as academic calendar dates. In this system, the chronons granularity is a set of milliseconds. Figure \ref{fig:granularity-graph-example} shows the complete granularity graph corresponding to this example. The transition between the chronons granularity and the seconds granularity is an example of a regular mapping. Regular mappings are represented by thin arrows in the visualisation of the graph. The transition between the days granularity and the months granularity is an example of an irregular mapping. In the graph visualisation, irregular mappings are represented by a bold arrow. Finally, the transition between the Gregorian calendar day granularity and the academic day granularity is an example of a congruent mapping. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. In the graph visualisation, congruent mappings are visualised as straight lines without arrow heads.

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.eps}
\caption{The granularity graph corresponding to example \ref{examplegrangraph}}
\label{fig:granularity-graph-example}
\end{figure}

@@

Several (groups of) operators have been proposed in order to compare temporal elements and model the temporal relationships between them. Allen~\cite{Allen83} most notably described such relationships between time intervals and as a special case, between instants. Figure \ref{fig:allen} shows the temporal relationships Allen discerned. Some proposals can be applied to both crisp and other time intervals~\cite{ohlbach2004},~\cite{nagypal2003},~\cite{schockaert08},~\cite{garrido2009}.

>>>>

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in~\cite{Bohlen1998}.

@@

\begin{example}
\label{examplegrangraph}
Consider a system that models both Gregorian calendar dates as well as academic calendar dates. In this system, the chronons granularity is a set of milliseconds. Figure~\ref{fig:granularity-graph-example} shows the complete granularity graph corresponding to this example. The transition between the chronons granularity and the seconds granularity is an example of a regular mapping. Regular mappings are represented by thin arrows in the visualisation of the graph. The transition between the days granularity and the months granularity is an example of an irregular mapping. In the graph visualisation, irregular mappings are represented by a bold arrow. Finally, the transition between the Gregorian calendar day granularity and the academic day granularity is an example of a congruent mapping. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. In the graph visualisation, congruent mappings are visualised as straight lines without arrow heads.

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.eps}
\caption{The granularity graph corresponding to example~\ref{examplegrangraph}}
\label{fig:granularity-graph-example}
\end{figure}

@@

Several (groups of) operators have been proposed in order to compare temporal elements and model the temporal relationships between them. Allen~\cite{Allen83} most notably described such relationships between time intervals and as a special case, between instants. Figure~\ref{fig:allen} shows the temporal relationships Allen discerned. Some proposals can be applied to both crisp and other time intervals~\cite{ohlbach2004},~\cite{nagypal2003},~\cite{schockaert08},~\cite{garrido2009}.
